Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2019-10-28 ms/manuscript.Rmd
Update MS

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern diversification: ca. 29% of all fern species are epiphytes vs. ca. 10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns. 
This raises the possibility that modern fern diversification was not shaped by the evolution of epiphytism *per se*, but rather by the exploitation of various niches within the canopy and on the forest floor.

@@

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a "filter", we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
However, the relative strength of this environmental filter may depend on overall abiotic conditions; for example, relative humidity (RH) and temperature also vary with elevation [@Cardelus2006]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities and reduced phylogenetic signal of relevant traits. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

@@

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestrial ferns for size-related traits (frond length, frond width, stipe length, and rhizome diameter) and pinna number (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Specific leaf area decreased with elevation, and was slightly lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval = `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "ntaxa", "p")`; \fig \@ref(fig:comm-div)). Species richness increased with elevation for both epiphytes and terrestrial communities (\fig \@ref(fig:comm-div)). Epiphytic communities were more diverse than terrestrial ones as measured by both MPD~func~ and MNTD~func~ (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "mpd.obs.z.func", "p")` and `r get_mod_result(t_test_print, "mntd.obs.z.func", "p")`, respectively). Functional diversity increased increased along the gradient for both terrestrial and epiphytic communities when measured with MPD~func~ (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "r")`, \pval = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "p")`), but not MNTD~func~. Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD~phy~ (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "mntd.obs.z.phy", "p")`) but not MPD~phy~ (\pval = `r get_mod_result(t_test_print, "mpd.obs.z.phy", "p")`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes exhibit reduced morphological diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (Fig. S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches.
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird’s nest ferns (e.g., *Asplenium nidus*), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird’s nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale has been reported in neotropical epiphtyic ferns as well [@Tausz2001; @Hietz2008]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz2008]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA) (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (\fig \@ref(fig:cwm-div)).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (i.e., thick, well-protected leaves) on Moorea: *Humata anderssonii* (a twig epiphyte), *Acrostichum aureum* (a terrestrial mangrove specialist), and *Paesia divaticatissima* (a terrestrial species occurring on high-elevation ridges) (Table S2).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

We did not find that gemmae production in gametophytes was linked with epiphytism. 
Gemmae are asexual propagules that allow gametophytes to persist over multiple growing seasons and attain large population sizes [@Farrar2008].
Gemmae production should be a useful trait in an epiphytic context because it may allow gametophytes to increase their surface area and hedge against population loss due to extreme drying events [@Farrar2008].
Gemmae can also function sexually by producing antherida in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001].
It is possible that some, but not all, epiphytic ferns rely on such strategies, and that they are not under similarly strong selection as other traits such as leaf size and gametophyte morphotype. 
Interestingly, when we classified 'widespread' species as those with gametophytes occurring above or below the minimum or maximum elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with widespread growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval = `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution).
Thus, it seems that gemmae may function in long-term gametophyte persistence, but are not required for epiphytic growth.

Presence of hairs and glands in gametophytes has very strong phylogenetic signal (\tab \@ref(tab:phylosig-binary)), and these traits are associated with certain taxonomic groups [@Stokey1951; @Nayar1971].
Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], and glands may affect rates of osmosis by modifying lipid concentrations [@Crow2011], we did not detect a correlation between hair or gland production and epiphytic growth.
Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established [@Johnson2012].
Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes.

### The role of environmental filtering in the canopy

Growth habit and temperature are clearly important factors in most of the community diversity metrics we studied, as shown by the full-subsets model analysis (\fig \ref{fig:heatmap}).
This is also supported by the linear elevation models (\figs \ref{fig:cwm-div}, \ref{fig:comm-div}), given that temperature shows a strong linear relationship with elevation (\fig \ref{fig:climate}).
It is somewhat puzzling that VPD did not contribute significantly to any models, as we hypothesized that this would be important for epiphytes that are frequently subjected to water stress.
One reason could be due to conditions at our study site: periods of prolonged rain result in the humidity sensor tip becoming wet, and it does not function normally again (i.e., shows 100% RH) until it dries out.
The raw data show several extended periods of 100% RH in terrestrial dataloggers at higher elevations.
Despite these possible inaccuracies however, epiphytic habitats are clearly drier than terrestrial (\fig \ref{fig:climate}).

We obtained partial support for the hypothesis that epiphytic fern communities would be more phylogenetically clustered than terrestrial ones; this was observed for MNTD, but not MPD (\fig \@ref(fig:comm-div)).
This makes sense given that most epiphytes have radiated within five major clades [@Schuettpelz2009]; each epiphytic community we sampled tends to include closely-related species from two or more epiphytic clades, but has a similar overall level of phylogentic diversity as terrestrial communities.
However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient.
It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends.
When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, @Nitta2017 detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation.
@Kluge2011 also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group.
They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)).
However, this result is dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1).
Furthermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)).
This suggests that epiphytic ferns are subject to environmental filtering for size-related traits, while simultaneously diversifying along other trait axes.
This could be maintained by niche partitioning within epiphytic habitats, which may be more diverse than terrestrial for a given area [@Rojas-Hernandez2018].
Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995].
The vertical stratification of epiphytic species has been well documented in ferns [\eg, @Parra2009; @WatkinsJr.2006a; @Acebey2017] and broader studies [\eg, @Kromer2007; @Zotz2007].
Future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

@@

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns.
Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits.
Finally, we observed drier conditions in the canopy relative to terrestrial habitats. 
Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simultaneously enabling diversification along other trait axes.
The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. 
Epiphytism itself is a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies.
Small size seems to be a prerequisite for epiphytic growth, as indicated both by our results and the global study of @Testo2018a.
Here, we have shown that size appears to be the exception, not the rule, when it comes to trait diversity of epiphytes: although epiphytes are universally small, our data suggest they are diverse in most other traits.
This emphasizes that fern epiphytes should not be treated as a monolithic group, but rather encompass a wide range of adaptive forms.
We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit.
Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.

@@

We thank members of the Davis Lab, N. Michele Holbrook, and Jonathan Losos for helpful discussion and comments on drafts. Suzanne Vinette, Saad Amer, Tristan Wang, and Tohei Theophilus helped with community surveys and measurement of traits. Staff at the University of California, Berkeley Richard B. Gump South Pacific Research Station, Moorea, French Polynesia, in particular Valentine Brotherson, Hinano Teavai-Murphy, Frank Murphy, and Neil Davies, provided support with obtaining research and collection permits, laboratory space, and logistics. Funding provided in part by National Science Foundation (Doctoral Dissertation Improvement Grant DEB-1311169 to JHN and CCD), Setup Funds from Harvard University to CCD, the American Society of Plant Taxonomists (Research Grant for Graduate Students to JHN), the Garden Club of America (Award in Tropical Botany to JHN), the Harvard University Herbaria (Fernald Fieldwork Fellowship to JHN), the Society of Systematic Biologists (Graduate Student Research Award to JHN), and the Systematics Association (Systematics Research Fund to JHN).

@@

J.H.N. conceived and carried out the study. R.T. collected data. J.H.N. performed the analyses. J.H.N., J.E.W., and C.C.D. wrote the manuscript.

@@

**\fig \@ref(fig:climate)**: Selected microclimatic variables of study plots along an elevational gradient from \SIrange{200}{1200}\meter on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from July 7, 2013 to July 5, 2014, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at ca. 2 m on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval \textless \num{.05}

**\fig \@ref(fig:pca)**: Principal components analysis (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. **(a, b)** PC loadings. **(c, d)** species scores. **(a)** and **(c)** are standard PCA, **(b)** and **(d)** are phylogenetic PCA. Epiphytes in green, terrestrial species in brown. Quantitative traits only, including only species with no missing observations for any traits ($\n = 104$ \si{\species}).

>>>>

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern diversification: \circa 29% of all fern species are epiphytes vs. \circa 10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns. 
This raises the possibility that modern fern diversification was not shaped by the evolution of epiphytism *per se*, but rather by the exploitation of various niches within the canopy and on the forest floor.

@@

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a "filter", we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities and reduced phylogenetic signal of relevant traits. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

@@

Community-weighted mean values (CWMs) were smaller for epiphytic than terrestrial ferns for size-related traits (frond length, frond width, stipe length, and rhizome diameter) and pinna number (\tval-test, all \pval < 0.05; \fig \@ref(fig:cwm-div)). Specific leaf area decreased with elevation, and was slightly lower for epiphytic relative to terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "sla", "r")`, \pval `r get_mod_result(model_summ_print, "sla", "p")`). Degree of lamina dissection decreased with elevation in epiphytes, but increased in terrestrial communities (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "dissection", "r")`, \pval = `r get_mod_result(model_summ_print, "dissection", "p")`). Frond length and width and rhizome diameter decreased slightly with elevation for terrestrial communities, while increasing slightly for epiphytic communities (\fig \@ref(fig:cwm-div)).

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "ntaxa", "p")`; \fig \@ref(fig:comm-div)). Species richness increased with elevation for both epiphytes and terrestrial communities (\fig \@ref(fig:comm-div)). Epiphytic communities were more diverse than terrestrial ones as measured by both MPD~func~ and MNTD~func~ (two-sided \tval-test, \pval `r get_mod_result(t_test_print, "mpd.obs.z.func", "p")` and `r get_mod_result(t_test_print, "mntd.obs.z.func", "p")`, respectively). Functional diversity increased along the gradient for both terrestrial and epiphytic communities when measured with MPD~func~ (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "mpd.obs.z.func", "r")`, \pval `r get_mod_result(model_summ_print, "mpd.obs.z.func", "p")`), but not MNTD~func~. Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD~phy~ (two-sided \tval-test, \pval `r get_mod_result(t_test_print, "mntd.obs.z.phy", "p")`) but not MPD~phy~ (\pval = `r get_mod_result(t_test_print, "mpd.obs.z.phy", "p")`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (\fig S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird’s nest ferns (e.g., *Asplenium nidus*), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird’s nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale has been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA) (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (\fig \@ref(fig:cwm-div)).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (i.e., thick, well-protected leaves) on Moorea: *Humata anderssonii* (a twig epiphyte), *Acrostichum aureum* (a terrestrial mangrove specialist), and *Paesia divaticatissima* (a terrestrial species occurring on high-elevation ridges) (Table S2).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

We did not find that gemmae production in gametophytes was linked with epiphytism. 
Rather, occurrence of gemmae strongly reflects phylogeny, appearing concentrated in filmy ferns (Hymenophyllaceae) and vittarioid ferns (\fig \@ref(fig:tree)).
Gemmae are asexual propagules occurring primarily in non-cordate gametophytes that allow gametophytes to persist over multiple growing seasons and attain large population sizes, and have been hypothesized to be associated with harsh environmental conditions [@Farrar2008].
In particular, gemmae production should be useful in an epiphytic context because they may allow gametophytes to increase their boundary layer and hedge against population loss during extreme drying [@Farrar2008].
Gemmae can also function sexually by producing antheridia in the presence of antheridiogen secreted by other gametophytes [@Emigh1977], which may confer a selective advantage in the epiphytic environment by enabling outcrossing [@Dassler2001].
As with our results summarized above, it is possible that the diversity of gemmae and non-gemmae producing epiphytic species reflects undetected niche variation in the canopy.
Interestingly, when we classified 'widespread' species as those with gametophytes occurring above or below the minimum or maximum elevational range of conspecific sporophytes using the data of @Nitta2017, we found that gemmae production is correlated with widespread growth (log likelihood dependent model = `r correlated_evo_test_gemmae$dependent.logL %>% round(1) %>% as.numeric`, log likelihood independent model = `r correlated_evo_test_gemmae$independent.logL %>% round(1) %>% as.numeric`, \pval = `r correlated_evo_test_gemmae$P %>% round(3) %>% as.numeric`, Pagel's test of correlated evolution), echoing a pattern observed in Japan of non-cordate gametophytes frequently occurring independently of their conspecific sporophytes [@Ebihara2013].

Finally, the presence of hairs and glands in gametophytes also exhibit very high phylogenetic signal (\ie, much greater than expected under Brownian motion; \tab \@ref(tab:phylosig-binary)), and these traits are concentrated in certain taxonomic groups [\eg, hairs, grammitid ferns; glands, Thelypteridaceae; @Stokey1951; @Nayar1971].
Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], and glands may affect rates of osmosis by modifying lipid concentrations [@Crow2011], we did not detect a correlation between hair or gland production and epiphytic growth.
Given their high phylogenetic signal, it is possible that these traits are unrelated to environment but rather reflect phylogenetically conserved developmental pathways [@Johnson2012].
Alternatively, the possibility remains that some epiphytic species rely on hairs or glands to prevent or slow drying depending on their niche within the canopy (\eg, hair-bearing grammitid gametophytes at mid to high elevation).

### The dual roles of the canopy as both a filter and promoter of diversity

Our results demonstrate that the canopy is far more dynamic with regard to fern epiphytism than has been previously appreciated. 
On the one hand, the canopy appears to have served as a strong barrier to entry, sorting species by size. 
Furthermore, the results of our phylogenetic community structure analyses partially support the hypothesis of decreased phylogenetic diversity in epiphytic communities due to filtering: we observed that epiphytic fern communities were more phylogenetically clustered for MNTD~phy~, but not MPD~phy~ (\fig \@ref(fig:comm-div)).
This makes sense given that most epiphytes have radiated within five major clades [@Schuettpelz2009]; each epiphytic community we sampled tends to include closely related species from two or more epiphytic clades, but exhibits a similar overall degree of phylogenetic diversity as terrestrial communities.

However, while the canopy clearly acts as a filter, we also find that it simultaneously functions as a promoter of diversity. 
This is evident in our analyses of MPD~func~ and MNTD~func~, which both indicate that epiphytic communities are more functionally diverse compared to terrestrial species (\fig \@ref(fig:comm-div)), thus contradicting our hypothesis of decreased functional diversity in epiphytic communities.
Additional analyses using different trait combinations support the role of non-size related traits driving this pattern (\fig S1).
We propose that the high functional diversity of epiphytic fern communities is maintained by previously undetected niche partitioning within epiphytic habitats as follows.
The canopy is generally drier than terrestrial habitats, exerting strong selection for small size. 
At the same time, microclimatic conditions in trees vary rapidly over short distances with height and distance from the trunk [@Hietz1995; @Cardelus2005], contributing to greater environmental variation per unit area than terrestrial habitats [@Rojas-Hernandez2018]. 
This vertical and horizontal diversity of niches is occupied by different fern species [@WatkinsJr.2006a; @Cardelus2007; @Parra2009; @Acebey2017].
Thus, when analyzed together, epiphytic fern communities encompass a wider range of morphological variation than terrestrial communities on average.
Assuming that the traits we measured reflect physiological strategies of epiphytes, this high morphological diversity likely derives from physiological adaptations to a variety of within-canopy niches. 
Clearly, fine-grained characterization of canopy niches and associated adaptations of epiphytic ferns occupying them is an important goal for future research.

### The primacy of growth habit in determining fern community composition as indicated by modelling

The primary role of growth habit, in combination with temperature, in structuring fern communities is supported by the full-subsets model analysis (\fig \ref{fig:heatmap}).
It is somewhat puzzling that VPD did not contribute significantly to any models, as we hypothesized that this would be important for epiphytes that are frequently subjected to water stress. 
Previous studies have shown marked variation in the air water potential across epiphytic habitats, with @WatkinsJr.2009 reporting a \circa \SI{40}{\mega\pascal} shift along the trunks of emergent canopy trees. 
Our epiphytic dataloggers only showed slightly greater VPD relative to those on the ground, with this difference largely dissipating at high elevations (\fig \ref{fig:climate}).
One reason for such low observed variation could be due to environmental conditions at our study site and placement of dataloggers: Moorea is a generally wet site overall (average annual rainfall \SI{325}{\centi\meter}), and we placed the epiphytic dataloggers on tree trunks, not in the higher emergent portions of the canopy.
Therefore, these measurements should be interpreted as reflecting the minimum amount of difference between terrestrial and canopy environments. 
Future studies characterizing within-canopy niches of epiphytic ferns should include placement of dataloggers at multiple sites throughout the canopy, which will allow for more detailed tests of the hypothesis that epiphytic fern diversity is maintained by niche partitioning.

@@

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth.
Our study provides one possible explanation for this finding. 
Here, we demonstrate that epiphytism is likely a more complex trait than previously realized, and almost certainly requires the evolution of multiple associated and dependent traits. 
We propose that the timing and order of the evolution of such traits are related to the complex underlying diversification dynamics observed in these studies. 
Future research investigating the tempo and mode of the evolution of epiphytism in ferns should seek to better partition this diversity when exploring hypotheses of exceptional diversification.

Small size appears to be a prerequisite for epiphytic growth, as indicated both by our results and by @Testo2018a.
However, we have shown that size appears to be the exception, not the rule, when it comes to trait diversity of epiphytes: although epiphytes are generally small, our data suggest they are diverse in numerous other traits. 
This emphasizes that fern epiphytes should not be treated as a single, static entity, but rather encompass a wide range of adaptive forms. 
Our results suggest new targets for future studies of fern ecology and evolution. 
We hope future global phylogenetic studies will build on our results by analyzing traits involved in epiphytic growth other than plant size (\eg, gametophyte morphotype, frond dissection, SLA, and hair and gland presence) to dissect how each may relate to diversification and growth habit. 
Additionally, future community ecological studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.

@@

We thank members of the Davis Lab, N. Michele Holbrook, and Jonathan Losos for helpful discussion and comments on drafts. Suzanne Vinette, Saad Amer, Tristan Wang, and Tohei Theophilus helped with community surveys and measurement of traits. Ravahere Taputuarai helped collect environmental data. Staff at the University of California, Berkeley Richard B. Gump South Pacific Research Station, Moorea, French Polynesia, in particular Valentine Brotherson, Hinano Teavai-Murphy, Frank Murphy, and Neil Davies, provided support with obtaining research and collection permits, laboratory space, and logistics. Funding provided in part by National Science Foundation (Doctoral Dissertation Improvement Grant DEB-1311169 to JHN and CCD), Setup Funds from Harvard University to CCD, the American Society of Plant Taxonomists (Research Grant for Graduate Students to JHN), the Garden Club of America (Award in Tropical Botany to JHN), the Harvard University Herbaria (Fernald Fieldwork Fellowship to JHN), the Society of Systematic Biologists (Graduate Student Research Award to JHN), and the Systematics Association (Systematics Research Fund to JHN).

@@

J.H.N. conceived and carried out the study and analyzed the data. J.H.N. and C.C.D. wrote the draft with editorial input from J.E.W.

@@

**\fig \@ref(fig:climate)**: Selected microclimatic variables of study plots along an elevational gradient from \SIrange{200}{1200}\meter on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from July 7, 2013 to July 5, 2014, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at \circa 2 m on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval \textless \num{.05}

**\fig \@ref(fig:pca)**: Principal components analysis (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. **(a, b)** PC loadings. **(c, d)** species scores. **(a)** and **(c)** are standard PCA, **(b)** and **(d)** are phylogenetic PCA. Epiphytes in green, terrestrial species in brown. Quantitative traits only, including only species with no missing observations for any traits ($\n = 104$ \si{\species}). Trait abbreviations as follows: dissection, degree of frond dissection; length, frond length; width, frond width; pinna, number of pinnae pairs; rhizome, rhizome diameter; stipe, stipe length; SLA, specific leaf area.
